Structural replication of nanoporous media using procedural noise

### 2018-05-30 ms.tex
More fixes

\begin{figure}
\includegraphics[width=.5\textwidth]{1d_noise.pdf}
\caption{1-dimensional simplex noise with 1 (yellow), 3 (red) and 30 (blue) octaves. Adding more octaves adds high frequency features, but it still follows the 1 octave base value.}
\label{fig:1dperlin}
\end{figure}

@@

\begin{align}
  \text{Delete}(\vec r) = 
  \begin{cases} 
    \text{True} & \text{if} \quad M_1(\vec r |\bar \theta) < \tau \\
    \text{False} & \text{if} \quad M_1(\vec r |\bar \theta) \geq \tau,
  \end{cases}
\end{align}
where we simply evaluate this function for every atom in the system.
The threshold $\tau$ is now added as a parameter to the model.

@@

Ideally, the model measure should pick up all geometrical properties that are
relevant for the physical study of interest. 
For instance, transport in porous media heavily depends on the pore size distribution and porosity\cite{kozeny1927uber, carman1937fluid},
whereas chemical reactons are sensitive to surface area\cite{coussy2011mechanics}. We have used $g(r)$ as defined and implemented in LAMMPS\cite{plimpton1995fast}.
In our cost function (the $\chi^2$ measure), we have ignored the low $r$ values of $g(r)$ since the pore geometry does not depend on the intra-molecular structure, but only the density variations at larger distances.

@@

As an additional validation, we also compute the porosity and surface area of the model since these are important quantities for fluid transport.
The method used for calculating porosity of a porous media is the same as in \cite{gelb1998characterization}.
Surface area is obtained by creating a voxelized grid of the system with values being the distance from each voxel center to the nearest atom.
This is now a scalar field and we can apply marching cubes\cite{lorensen1987marching} to triangulate an isosurface of the distance field.
This gives a trianglated representation of the porous media defined by a set of atomic positions.
As discussed in \cite{gelb1998characterization}, the definition of surface area is ill-defined in
atomistic simulations, because a clear definition of pore space is difficult.
We have chosen to define the pore space as the set of points being more than \SI{2}{\angstrom} to any atom.
This is large enough to contribute to a dense pore matrix, but small enough to have most of the available pore space defined as such. 
The important note here is that the same definition of both surface and porosity is being used in all systems that are compared, physical simulations, procedural methods or data.

@@

The posterior probability distribution is defined as the probability of the parameters
$\bar \theta$ given the evidence $d$, or $P(\bar \theta | d)$. Bayes' theorem states that 
\begin{equation}
P(\bar \theta | d) = \frac{P(d | \bar \theta)P(\bar \theta)}{P(d)} \propto P(d | \bar \theta)P(\bar \theta),
\end{equation}
where $P(d | \bar \theta)$ is called the Likelihood $\mathcal L$ (the probability of
evidence given parameters) and $P(\bar \theta)$ is the prior (the beliefs about the
quantity before some evidence is taken into account). $P(d)$ is called the evidence,
and is the probability of the observations (here assumed to be 1). Prior parameters
are restricted to freely roam within a minimum and maximum value, so we can therefore
assume that the posterior distribution is proportional to the likelihood, or
$P(d | \bar \theta) \propto \mathcal L$. In other words, to calculate the posterior
probability distribution and obtaining the best fit model parameters, we need only
to estimate the likelihood function, which is connected to the $\chi^2$ by 
\begin{equation}
\mathcal L \propto e^{-\chi^2}.
\end{equation}

In order to estimate whether two 3D data sets are statistically equivalent up to
the model measure $g(r)$, we calculate the $\chi^2$ between the two one-dimensional
RDFs. Pearson's $\chi^2$ is defined as 
\begin{equation}
  \chi^2 = \sum_r \frac{ \Big(g_d(r) - g_{M(\bar \theta)}(r) \Big)^2}{g_{M(\bar \theta)}(r)^2},
\end{equation}
so for low values of $\chi^2$, the better the fit between data and model, up to
$g(r)$. For a single-parameter model, we could simply just iterate over the lone
parameter to obtain the minimum $\chi^2$ (where the best-fit parameter value resides),
however, with increased number of parameters comes increased computational cost.
If we divide the parameter space into $N$ bins, the number of $\chi^2$ computations
scales as $N^d$, where $d$ is the number of model parameters. To be able to build
the full landscape of the posterior probability distribution, we need to use the Monte Carlo Markov chain (MCMC) method .

We are now interested in maximising the likelihood $\mathcal L$ as opposed to
minimising the $\chi^2$, with the goal of mapping out the full $d$-dimensional
likelihood space for the model parameters. We build the $d$-dimensional likelihood
by letting random walkers traverse the parameter space guided by the MCMC test:
for each step $M(\bar \theta_{n})$ in parameter space, we calculate the likelihood
$\mathcal L_n$ for this particular parameter configuration. Then, for the next
step, we create a new set of parameters $M(\bar \theta_{n+1})$ by choosing new
parameters randomly from a $d$-dimensional sphere around the previous parameter
composition. The model also selects a new random seed. We calculate the proposed
likelihood $\mathcal L_{n+1}$ at this new step and perform the MCMC test: accept
the new step if $\frac{\mathcal L_n}{\mathcal L_n+1} \ge 1$, or reject if the value
is lower than a uniform random value $\in [0,1]$.

Finally, we obtain the posterior $P(\bar \theta | d) \propto \mathcal L$ by building
a $d$-dimensional histogram of the likelihood using the stored positions of the parameters.
For a sufficiently large amount of steps, the MCMC algorithm ensures that the posterior
converges to the likelihood, as long at the surface does not contain unreachable areas
such as discontinuous patches or an infinite amount of maxima (as is the case for the
function $f(x)=sin(\frac{1}{x}$). Luckily, as we shall see, the likelihood surface for
our models tend to turn out with a single, well-defined maximum.

\section{Results}
We now turn our attention to applying the MCMC framework on two kinds of data.
First, we wish to test and verify the code by performing a full data analysis on
mock data with known input parameters. This enables us to investigate the posterior
distribution, and determine whether any potential problems could arise with the
likelihood landscape, such as random walkers trapped in local minima. In addition,
we will obtain an intuition on the width of the estimated parameter distributions.
But most importantly, we will have verified that our likelihood code is working correctly
if we successfully reproduce the original input parameters. We then continue with a
full analysis of simulated SiO$_2$ data, and see whether the resulting best-fit
parameter data set is consistent with the SiO$_2$ data set.

\subsection{Model verification: Parameter estimation of mock data}
\begin{figure}[htb!]
\includegraphics[width=.45\textwidth]{model_test.png}
\caption{Simulated mock data used for the initial model testing analysis. Input parameters are $\nu=0.03$,$\tau=-0.02$ and $\psi=0.6$.}
\label{fig:mockdata}
\end{figure}
In order to verify our likelihood code, we perform a model verification test
on a mock data set with known input values based on equation \ref{eq:noisemodel1}.
In this example, we only use standard simplex noise with three free parameters:
the persistence $\psi$, threshold $\tau$ and scale $\nu$. We simulate a data set
of size $\SI{159} {\angstrom}^3$ with $\sim 255 000$ atoms, and perform the cut-off
given by input parameters $\Omega=3$, $\Psi = 0.6$, $\tau=-0.2$ and $\nu=0.03$.
The resulting 3D structure is depicted in figure \ref{fig:mockdata}.

We continue by executing a full MCMC analysis on the mock data by using a
parallelized code written in C++ running on a 48 core cluster. About 48 hours
was needed to produce about 300 000 samples. The results from this analysis
can be seen in figure \ref{fig:mockdataresults_pts}, where the upper panel
depicts the marginalised 1D posteriors while the lower shows the marginalised
2D posteriors. The centre and left image shows that the input parameters of
$\tau=-0.2$(threshold), $\psi=0.6$(persistence) and $\nu=0.03$ are all reproduced
well within $1 \sigma$. Note that there seems to be no problems with the posterior
distribution in terms of local minima or multiple maxima. In addition, the results
show that all the parameters are correlated with $r \sim 0.85$, while the width of
the marginalised distributions vary by $\sigma_\tau = 0.02 \sim 10\%$, $\sigma_\Psi = 0.062 \sim 20\%$ and $\sigma_\nu = 0.01 \sim 33\%$.

@@

\label{fig:porous_results1}
\end{figure*}
Having verified that the likelihood code yields correct results when analysing mock data, we turn to 
performing a parameter fitting analysis on realistic simulated nanoporous media.
Using LAMMPS \cite{plimpton1995fast} and the Vashishta-potential \cite{vashishta1990interaction},
we simulate system consisting of $14^3$ betacristobalite unit cells of SiO$_2$, corresponding
to system size of approximately \SI{100}{\angstrom}. 
The temperature is then increased to \SI{4500}{\kelvin} using NVT integration. 
After an initial period, we expand the box to \SI{159}{\angstrom} to achieve a porosity $\phi=0.75$. 
Then, the system is quenched by setting the temperature to \SI{300}{\kelvin},
resulting in the nanoporous media on the right side in figure \ref{fig:porous_vs_model}. 
By choosing a different seed for the initial velocities, it is possible to produce
statistically equal samples but with random realizations.

With the model described in equation \ref{eq:noisemodel1}, we perform a full likelihood
analysis of the data set with the three free parameters $\tau$, $\nu$ and $\psi$ using $g(r)$ as the measure. 
The resulting marginalised posteriors are depicted in figure \ref{fig:porous_results1},
where the best-fit parameters $\tau=-0.2$, $\psi=0.8$ and $\nu=0.3$ is extracted from the full 3D likelihood. 
We also display $g(r)$ for the input data set, a best-fit parameter data set and a
bad-fit data set in figure \ref{fig:gofr1}, to illustrate the sensitivity of $g(r)$ given the model parameters. 
Note that we are not able to accurately reproduce the structures of the SiO$_2$ $g(r)$ with
our current model, but hope that we during future work will be able to produce another model
that provides a better fit with this measure. 
\todo{skriv om hvorfor vi ikke plukker opp alle features?}

Note that the surface atoms are not in chemical equilibrium since some neighbours of a surface atom might have been removed. 
In order to use the generated geometry in molecular dynamics, one would need either noble gas
atoms which may be stable, or to apply passivation step.

>>>>

\begin{figure}
\includegraphics[width=.5\textwidth]{1d_noise.pdf}
\caption{1-dimensional simplex noise with 1 (yellow), 3 (red) and 30 (blue) octaves. Adding more octaves adds high frequency features, but it still follows the 1 octave base signal.}
\label{fig:1dperlin}
\end{figure}

@@

\begin{align}
  \text{Delete}(\vec r) = 
  \begin{cases} 
    \text{True} & \text{if} \quad M_1(\vec r |\bar \theta) < \tau \\
    \text{False} & \text{if} \quad M_1(\vec r |\bar \theta) \geq \tau,
  \end{cases}
  \label{eq:delete_threshold}
\end{align}
where we simply evaluate this function for every atom in the system.
The threshold $\tau$ is now added as a parameter to the model.

@@

Ideally, the model measure should pick up all geometrical properties that are
relevant for the physical study of interest. 
For instance, transport in porous media heavily depends on the pore size distribution and porosity\cite{kozeny1927uber, carman1937fluid},
whereas chemical reactons are sensitive to surface area\cite{coussy2011mechanics}. We have used $g(r)$ as defined and implemented in LAMMPS\cite{plimpton1995fast}.
In our cost function (the $\chi^2$ measure), we have ignored the values of $g(r)$ for $r<$\SI{3}{\angstrom} since the pore geometry does not depend on the intra-molecular structure, but only the density variations at larger distances.
The cutoff value of \SI{3}{\angstrom} was chosen since the peak of $g(r)$ is at \SI{2.9}{\angstrom} for Si-O\cite{vashishta1990interaction}, and has .

@@

As an additional validation, we also compute the porosity and surface area of the model since these are important quantities for fluid transport.
The method used for calculating porosity of a porous media is the same as in \cite{gelb1998characterization}.
Surface area is obtained by creating a voxelized grid of the system with values being the distance from each voxel center to the nearest atom.
This is now a scalar field and we can apply marching cubes\cite{lorensen1987marching} to triangulate an isosurface of the distance field.
This gives a triangulated representation of the porous media defined by a set of atomic positions.
As discussed in \cite{gelb1998characterization}, the definition of surface area is ill-defined in
atomistic simulations, because a clear definition of pore space is difficult.
We have chosen to define the pore space as the set of points being more than \SI{2}{\angstrom} to any atom.
This is large enough to contribute to a dense pore matrix, but small enough to have most of the available pore space defined as such. 
The important note here is that the same definition of both surface and porosity is being used in all systems that are compared, physical simulations, procedural methods or data.

@@

The posterior probability distribution is defined as the probability of the parameters
$\bar \theta$ given the evidence $d$, or $P(\bar \theta | d)$. Bayes' theorem states that 
\begin{equation}
P(\bar \theta | d) = \frac{P(d | \bar \theta)P(\bar \theta)}{P(d)} \propto P(d | \bar \theta)P(\bar \theta),
\end{equation}
where $P(d | \bar \theta)$ is called the Likelihood $\mathcal L$ (the probability of
evidence given parameters) and $P(\bar \theta)$ is the prior (the beliefs about the
quantity before some evidence is taken into account). $P(d)$ is called the evidence,
and is the probability of the observations (here assumed to be 1). 
To calculate the posterior probability distribution and obtain the best fit model parameters, we need only
to estimate the likelihood function, which is connected to the $\chi^2$\todo{Why do we have an exponential dependence of $\chi^2$ here?} by 
\begin{equation}
\mathcal L \propto e^{-\chi^2}.
\end{equation}

In order to estimate whether two 3D data sets are statistically equivalent up to
the model measure $g(r)$, we calculate the $\chi^2$ between the two one-dimensional
RDFs. We calculate $g(r)$ using a histogram over spherical shells of distance $r_i$ to the center for bin $i$. Pearson's $\chi^2$ is then defined as 
\begin{equation}
  \chi^2 = \sum_{i} \frac{ \Big(g_d(r_i) - g_{M(\bar \theta)}(r_i) \Big)^2}{g_{M(\bar \theta)}(r_i)^2},
\end{equation}
which means that for low values of $\chi^2$, the better the fit between data and model, up to $g(r)$.

The difficult task is now to find which set of parameters $\bar\theta$ that maximizes the likelihood $\mathcal L$.
For a single-parameter model with parameter $\psi$, we could simply just discretize the interval $[\psi_\text{min}, \psi_\text{max}]$ into $N$ values,
and loop over all parameter to obtain the minimum $\chi^2$.
However, each parameter is an extra dimension in the parameter space, with increases the computational cost substantially.
If each parameter is discretized into into $N$ values, the full number of $\chi^2$ computations
scales as $N^d$, where $d$ is dimension of the parameter space - i.e. the number of model parameters. To be able to build
the full landscape of the posterior probability distribution, we have chosen to use the Monte Carlo Markov chain (MCMC) method.

We are now interested in maximising the likelihood $\mathcal L$ as opposed to
minimising the $\chi^2$, with the goal of mapping out the full $d$-dimensional
likelihood space for the model parameters. We build the $d$-dimensional likelihood
by letting random walkers traverse the parameter space guided by the MCMC test:
for each step $M(\bar \theta_{n})$ in parameter space, we calculate the likelihood
$\mathcal L_n$ for this particular parameter configuration. Then, for the next
step, we create a new set of parameters $M(\bar \theta_{n+1})$ by choosing new
parameters randomly from a $d$-dimensional sphere around the previous parameter
composition. We calculate the proposed
likelihood $\mathcal L_{n+1}$ at this new step and perform the MCMC test: accept
the new step if $\frac{\mathcal L_n}{\mathcal L_n+1} \ge 1$, or reject if the value
is lower than a uniform random value $\in [0,1]$.

Finally, we obtain the posterior $P(\bar \theta | d) \propto \mathcal L$ by building
a $d$-dimensional histogram of the likelihood using the stored positions of the parameters.
For a sufficiently large amount of steps, the MCMC algorithm ensures that the posterior
converges to the likelihood, as long at the surface does not contain unreachable areas
such as discontinuous patches or an infinite amount of maxima (as is the case for the
function $f(x)=\sin(\frac{1}{x}$)). Luckily, as we shall see, the likelihood surface for
our models tend to turn out with a single, well-defined maximum.

\section{Results}
We now turn our attention to applying the MCMC framework on two kinds of data.
First, we wish to test and verify the code by performing a full data analysis on
mock data with known input parameters. This enables us to investigate the posterior
distribution, and determine whether any potential problems could arise with the
likelihood landscape, such as random walkers trapped in local minima. In addition,
we will obtain an intuition on the width of the estimated parameter distributions.
But most importantly, we will have verified that our likelihood code is working correctly
if we successfully reproduce the original input parameters. We then continue with a
full analysis of physical system simulated with molecular dynamics, and seek parameters that reproduces $g(r)$ of this system.
\todo{Figure 5 is the same as left in fig 9.}
\subsection{Model verification: Parameter estimation of mock data}
\begin{figure}[htb!]
\includegraphics[width=.45\textwidth]{model_test.png}
\caption{Simulated mock data used for the initial model testing analysis. Input parameters are $\nu=0.03$,$\tau=-0.02$ and $\psi=0.6$.}
\label{fig:mockdata}
\end{figure}
In order to verify our likelihood code, we perform a model verification test
on a mock data set with known input values based on equation \ref{eq:noisemodel1}.
In this example, we only use standard simplex noise with three free parameters:
the persistence $\psi$, threshold $\tau$ and scale $\nu$.

The physical system is a periodic box with amorphous SiO$_2$ of size $\SI{159} {\angstrom}^3$ (255552 atoms), created with a heat-and-quench method.
A nanoporous system is created by removing atoms from the system accordin to equation \eqref{eq:delete_threshold} with parameters $\Omega=3$, $\Psi = 0.6$, $\nu=0.03$ and $\tau=-0.2$.
The resulting 3D structure is depicted in figure \ref{fig:mockdata}.

We continue by executing a full MCMC analysis on the mock data by using a
parallelized code written in C++ running on a 48 core cluster. About 48 hours
was needed to produce about 300000 samples. The results from this analysis
can be seen in figure \ref{fig:mockdataresults_pts}, where the upper panel
depicts the marginalised 1D posteriors while the lower shows the marginalised
2D posteriors. The centre and left image shows that the input parameters of
$\tau=-0.2$(threshold), $\psi=0.6$(persistence) and $\nu=0.03$ are all reproduced
well within $1 \ \sigma$ ($\Omega$ was kept fixed equal to 3). Note that there seems to be no problems with the posterior
distribution in terms of local minima or multiple maxima. In addition, the results
show that all the parameters are correlated with $r \sim 0.85$, while the width of
the marginalised distributions vary by $\sigma_\tau = 0.02 \sim 10\%$, $\sigma_\Psi = 0.062 \sim 20\%$ and $\sigma_\nu = 0.01 \sim 33\%$.

@@

\label{fig:porous_results1}
\end{figure*}
Having verified that the likelihood code yields satisfying results when analysing mock data, we turn to 
performing a parameter fitting analysis on realistic simulated nanoporous media.
Using LAMMPS\cite{plimpton1995fast} and the Vashishta-potential\cite{vashishta1990interaction},
we simulate system consisting of $14^3$ stretched\footnote{The nanoporous material will be amorphous, so we have stretched the initial crystal to match the density of amorphous silica.} betacristobalite unit cells of SiO$_2$, corresponding
to system size of approximately \SI{100}{\angstrom}. 
The system is then melted at \SI{4500}{\kelvin} using the Nose-Hoover thermostat.
After a few nanoseconds, the system box is slowly expanded to \SI{159}{\angstrom} to achieve a porosity of $\phi=0.75$ while keeping the thermostat on to keep the temperature constant.
Then, the system is quenched by setting the target thermostat temperature to \SI{300}{\kelvin},
resulting in the nanoporous media on the right side in figure \ref{fig:porous_vs_model}. 
By repeating this for different initial velocities, we can produce statistically similar samples with random realizations.

With the model described in equation \ref{eq:noisemodel1}, we perform a full likelihood
analysis of the data set with the three free parameters $\tau$, $\nu$ and $\psi$ using $g(r)$ as the measure. 
The resulting marginalised posteriors are depicted in figure \ref{fig:porous_results1},
where the best-fit parameters $\tau=-0.2$, $\psi=0.9$ and $\nu=0.03$ is extracted from the full 3D likelihood. 
We also display $g(r)$ for the input data set, a best-fit parameter data set and a
bad-fit data set in figure \ref{fig:gofr1}, to illustrate the sensitivity of $g(r)$ given the model parameters.
$\chi^2$ is only calculated for $r\in (3,15)$ since the nanoporous structure signal is found at larger $r$.
The main features of $g(r)$ from the data is picked up by our best-fit model and the bad fit (which has very high porosity) has slower convergence.
We see that we don't accurately reproduce all structures of the SiO$_2$ $g(r)$ with
our current model. For instance, the long tail is consistantly slightly lower in our model, and some of the bumps have details that differ from what our model gives.

\todo{skriv om hvorfor vi ikke plukker opp alle features?}

Also note that the surface atoms are not in chemical equilibrium since some neighbours of a surface atom might have been removed. 
In order to use the generated geometry in molecular dynamics, one would need either noble gas
atoms which may be stable, or to apply passivation step.
